---
title: "Lab Title"
author: "Erik Blomberg"
date: "`r format(Sys.time(), '%m/%d/%Y')`"
header-includes:
  - \usepackage{setspace}\onehalfspacing
  - \setlength\parindent{12pt}
fontsize: 12pt
output:
  pdf_document:
    toc: yes
    df_print: paged
    number_sections: yes
    latex_engine: xelatex
    
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo=FALSE}
library(knitr)
opts_chunk$set(tidy.opts=list(width.cutoff=60),tidy=TRUE)
```

```{css, echo=FALSE}
.scroll-200 {
  max-height: 200px;
  max-width: 100%;
  overflow-x: auto
  overflow-y: auto;
  background-color: inherit;
}
```

```{r echo=FALSE, include=FALSE}
#install.packages("RMark")
library(RMark)
#install.packages("bookdown")
library(bookdown)
library(ggplot2)
#install.packages('curl')
library(curl)
#install.packages("kableExtra")
library(kableExtra)

cleanup(ask=FALSE)

```

# Lab Overview

Section for a few sentences of generic setup

## Learning objectives

Typically 3 or 4 major learning objectives

# Lab Content

This will contain the meat of the lab including background materials and step by step

## Subheadings 

Typically both Section headings and subheadings are used

Equation example:

\begin{equation}
 \label{eq:1}
  \frac{\hat{N}}{M}=\frac{C}{m}
\end{equation}

Code example: (cheatsheet at https://rstudio.com/wp-content/uploads/2015/02/rmarkdown-cheatsheet.pdf)

```{r eval=FALSE, echo=TRUE, warning=FALSE, error=FALSE, message=FALSE, tidy=TRUE, results='markup'}

Species<- sample(c("White-tailed Deer", "Wild Hog", "Wooly Mammoth", "White-footed Mouse", "Wallaby",
                   "Wandering Albatross", "White Pelican", "Wilson's Warbler", "Whooping Crane", "Wood Duck",
                   "Wood Frog", "Whiptail", "Western Pond Turtle", "Western Diamondback", "Wood Turtle",
                   "White Bass", "Whale Shark", "Wahoo", "Walleye", "Wobbegong"),1)
  
Species 

```

# The Three Big Things You Should Have Learned Today

**FIRST** 

**SECOND** 

**THIRD** 


# Lab X Assignment


# Lab Appendix

## Quick reference for important R commands in this lab

## For more information

## Bibliography

**Term Highlighted in Lab** - Description


